---
title: "Hello, Quarto"
format: html
---

# R Package jmpwashdata

For this analysis we will use the **jmpwashdata** R Package. The package contains all data compiled by the [WHO/UNICEF Joint Monitoring Programme (JMP)](https://washdata.org).

```{r}
#| label: load-packages
#| message: false

library(jmpwashdata)
library(tidyverse)
library(gt)
library(ggthemes)
```

# World Bank income groups

We will also use the World Bank income classification for 218 countries. This data was downloaded and stored as an XLSX file using an R script in `src`.

```{r}

income_groups_df <- read_rds("data/wb-income-groups.rds")

```

# Basic Sanitation & GDP

Data for the most recent year, basic sanitation in urban areas, calculate urban population, and join income groups.

```{r}
# Perform data manipulation operations on the jmp_wld_sanitation data frame
jmp_wld_sanitation_gdp_income <- jmp_wld_sanitation |> 
  # Filter the rows where the year column is equal to the maximum year value
  filter(year == max(year)) |> 
  # Select the columns from name to prop_u and the san_bas_u column
  select(name:prop_u, san_bas_u) |> 
  # Create a new column named pop_u
  mutate(pop_u = pop_n * 1000 * prop_u / 100) |> 
  # Drop the pop_n and prop_u columns
  select(-pop_n, -prop_u) |> 
  # Perform a left join with the income_groups_df data frame
  left_join(income_groups_df) |> 
  # Drop the rows that have missing values in the san_bas_u & income_group cols
  drop_na(san_bas_u, income_group)

```

# Basic Sanitation Uganda

```{r}

# Create a vector of color codes
color_scale_sanitation <- c("#8cce8f", "#fff381", "#ffda5a", "#ffbc02")

# Create a vector of sanitation indicators
fct_sanitation <- c("basic", "limited", "unimproved", "open defecation")

# Perform data manipulation operations on the jmp_wld_sanitation data frame
jmp_uga_sanitation <- jmp_wld_sanitation |> 
  # Filter the rows where the iso3 column is equal to "UGA" and the year column 
  # is equal to 2000 or 2020
  filter(iso3 == "UGA") |> 
  filter(year == 2000 | year == 2020) |> 
  # Select the name, iso3, year, and columns from san_bas_n to san_od_n
  select(name, iso3, year, san_bas_n:san_od_n) |> 
  # Reshape the data frame from wide to long format
  pivot_longer(cols = san_bas_n:san_od_n, 
               names_to = "indicator", 
               values_to = "percent") |> 
  # Rename the indicator column based on the values of the san_bas_n to san_od_n
  # columns
  mutate(indicator = case_when(
    indicator == "san_bas_n" ~ "basic",
    indicator == "san_lim_n" ~ "limited",
    indicator == "san_unimp_n" ~ "unimproved",
    indicator == "san_od_n" ~ "open defecation"
  )) |> 
  # Convert the indicator column to a factor with levels specified by the 
  # fct_sanitation vector
  mutate(indicator = factor(indicator, level = fct_sanitation))

```

# Income

Below is a box- and jitterplot of countries with percentages of populations with access to basic sanitation in 2020 grouped by income classifications.

```{r}
#| label: san-bas-u-income
#| fig-alt: |
#|   A boxplot with jittered points overlaid. The x-axis represents income 
#|   groups, while the y-axis represents the percentage of the population with 
#|   access to basic sanitation. There is a positive, linear association
#|   increases.

ggplot(data = jmp_wld_sanitation_gdp_income,
       mapping = aes(x = income_group, 
                     y = san_bas_u, 
                     color = income_group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 4, alpha = 0.5) +
  labs(x = NULL, y = "percent") +
  scale_color_colorblind() +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none")
```

# Regions

The table below shows urban sanitation indicators for global regions in 2020.

```{r}
#| label: reg-sanitation

jmp_reg_sanitation |> 
  filter(year == max(year)) |> 
  filter(!str_detect(region, "income")) |> 
  select(region, san_bas_u:san_od_u) |> 
  drop_na() |> 
  gt(rowname_col = "region") |> 
  cols_label(
    san_bas_u = md("**basic**"),
    san_lim_u = md("**limited**"),
    san_unimp_u = md("**unimproved**"),
    san_od_u = md("**open defecation**")
  ) |> 
  fmt_percent(columns = san_bas_u:san_od_u,
              decimals = 0, 
              scale_values = FALSE) 

```

# Uganda

The figure below shows the sanitation ladder for Uganda.

```{r}
#| label: san-uga
#| fig-cap: Sanitation indicators for Uganda on a national level.

ggplot(data = jmp_uga_sanitation, 
       mapping = aes(x = year, 
                     y = percent, 
                     fill = indicator)) +
  geom_area() +
  labs(title = "Uganda: sanitation ladder (national)",
       x = NULL, y = "percent", fill = "indicators") +
  scale_fill_manual(values = color_scale_sanitation) +
  scale_x_continuous(breaks = c(2000, 2020)) +
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor = element_blank()) 

```
